---
title: "MirandaDDS8501-8"
author: "Daniel Miranda Ruiz"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Upload Libraries

```{r}
library(ggplot2)            # Load ggplot2 library
library(scales)             # Load scales library
library(tidyverse)          # Load tidyverse library
library(corrplot)     # Correlation visualization
library(caret)        # Modeling prep
library(psych)        # Descriptive statistics
library(Amelia)
library(dplyr)
```

## Get working directory

```{r}
getwd()
```

## See Files in directory

```{r}
list.files() # Look at Folder Files
```

## Load Dataset Bankrupcy companies

```{r}
american_banktupcy_dataset <- read.csv("american_bankruptcy.csv")
```

## Dataset Sample

```{r}
head(american_banktupcy_dataset)

```

## Variable Renaming for data undersatnding

```{r}
american_banktupcy_dataset <- american_banktupcy_dataset %>%
  rename(
    current_assets = X1,
    cost_of_goods_sold = X2,
    depreciation_amortization = X3,
    ebitda = X4,
    inventory = X5,
    net_income = X6,
    total_receivables = X7,
    market_value = X8,
    net_sales = X9,
    total_assets = X10,
    total_long_term_debt = X11,
    ebit = X12,
    gross_profit = X13,
    total_current_liabilities = X14,
    retained_earnings = X15,
    total_revenue = X16,
    total_liabilities = X17,
    total_operating_expenses = X18
  )
```

## Confirming rename

```{r}
colnames(american_banktupcy_dataset)
```

## Dataset Summary

```{r}

cat("\n\n===== DATA STRUCTURE =====================================================================\n\n")

str(american_banktupcy_dataset)

cat("\n\n===== DATA STRUCTURE =====================================================================\n\n")

glimpse(american_banktupcy_dataset)

cat("\n\n===== DATA STRUCTURE =====================================================================\n\n")

sapply(american_banktupcy_dataset, class)

cat("\n\n===== Summary Statistics =====================================================================\n\n")

summary(american_banktupcy_dataset)

cat("\n\n===== Summary Statistics =====================================================================\n\n")
 
library(dplyr)
library(tidyr)
library(moments)

numeric_data <- american_banktupcy_dataset %>%
  select(where(is.numeric))

summary_table <- numeric_data %>%
  summarise(across(everything(),
                   list(
                     Mean = ~mean(., na.rm = TRUE),
                     SD = ~sd(., na.rm = TRUE),
                     Min = ~min(., na.rm = TRUE),
                     Q1 = ~quantile(., 0.25, na.rm = TRUE),
                     Median = ~median(., na.rm = TRUE),
                     Q3 = ~quantile(., 0.75, na.rm = TRUE),
                     Max = ~max(., na.rm = TRUE),
                     Skew = ~skewness(., na.rm = TRUE),
                     Kurtosis = ~kurtosis(., na.rm = TRUE)
                   )))

summary_table <- summary_table %>%
  pivot_longer(everything(),
               names_to = c("Variable", ".value"),
               names_sep = "_") %>%
  mutate(across(where(is.numeric), round, 2))

summary_table
```

## Frequency Table for status_label

```{r}
table(american_banktupcy_dataset$status_label)

library(dplyr)

distribution_table <- american_banktupcy_dataset %>%
  count(status_label) %>%
  mutate(
    Percent = round(n / sum(n) * 100, 2)
  ) %>%
  rename(
    Status = status_label,
    Total = n
  )

distribution_table
```

## Frequencies for year
```{r}
year_distribution <- american_banktupcy_dataset %>%
  count(year) %>%
  mutate(
    Percent = round(n / sum(n) * 100, 2)
  ) %>%
  rename(
    Year = year,
    Total = n
  )

year_distribution
```


## Set status_label as factor

```{r}
american_banktupcy_dataset$status_label <- factor(american_banktupcy_dataset$status_label,
                                     levels = c("alive", "failed"))
```

## Finding Missing Data

```{r}
colSums(is.na(american_banktupcy_dataset))
```

## Missingness Map of Data Identifying Missing Data by Row and Column Using Amelia

```{r}
missmap(american_banktupcy_dataset,
  main   = "Missingness Map: Dataset",
  legend = TRUE,
  col    = c("red", "blue") 
)
```

## CORRELATION MATRIX for outliers

```{r}
numeric_vars <- american_banktupcy_dataset %>%
  select(where(is.numeric))

cor_matrix <- cor(numeric_vars, use = "complete.obs")

round(cor_matrix, 2)

library(corrplot)

corrplot(cor_matrix,
         method = "color",
         type = "upper",
         order = "hclust",
         tl.cex = 0.7,
         tl.col = "black")
```

# SCATTERPLOT MATRIX

```{r}

library(tidyverse)

american_banktupcy_dataset %>%
  select(where(is.numeric)) %>%
  pivot_longer(cols = everything(),
               names_to = "variable",
               values_to = "value") %>%
  ggplot(aes(x = "", y = value)) +
  geom_boxplot() +
  facet_wrap(~variable, scales = "free") +
  theme_minimal()
```

# Total Assets Distribution

```{r}
ggplot(american_banktupcy_dataset, aes(x = total_assets)) +
  geom_histogram(bins = 50) +
  scale_x_log10() +
  theme_minimal()
```

# Market Value Distribution

```{r}
ggplot(american_banktupcy_dataset, aes(x = market_value)) +
  geom_histogram(bins = 50) +
  scale_x_log10() +
  theme_minimal()
```

# Year vs count of alive and failed companies

```{r}
library(dplyr)
library(ggplot2)

year_summary <- american_banktupcy_dataset %>%
  group_by(year, status_label) %>%
  summarise(count = n(), .groups = "drop")

ggplot(year_summary, aes(x = year, y = count, color = status_label)) +
  geom_line(size = 1) +
  geom_point() +
  labs(title = "Trend of Active vs Bankrupt Firms Over Time",
       x = "Year",
       y = "Number of Firms",
       color = "Company Status") +
  theme_minimal()

ggplot(year_summary, aes(x = factor(year), y = count, fill = status_label)) +
  geom_bar(stat = "identity") +
  labs(title = "Distribution of Firm Status by Year",
       x = "Year",
       y = "Number of Firms") +
  theme_minimal()
```

## Net Income distributions

```{r}

ggplot(american_banktupcy_dataset, aes(x = total_liabilities,
                          y = net_income,
                          color = status_label)) +
  geom_point(alpha = 0.5) +
  scale_x_log10() +
  scale_y_log10() +
  facet_wrap(~ year) +
  theme_minimal() +
  labs(title = "Net Income vs Total Liabilities by Year (Log Scale)",
       x = "Total Liabilities (log scale)",
       y = "Net Income (log scale)",
       color = "Status")


```

## Density Plot for total Liabilities by status

```{r}
ggplot(american_banktupcy_dataset, aes(x = total_liabilities,
                          fill = status_label)) +
  geom_density(alpha = 0.4) +
  scale_x_log10() +
  theme_minimal() +
  labs(title = "Distribution of Total Liabilities by Firm Status",
       x = "Total Liabilities (log scale)",
       fill = "Status")
```

## Density Plot for total net income by status

```{r}
ggplot(american_banktupcy_dataset, aes(x = net_income,
                          fill = status_label)) +
  geom_density(alpha = 0.4) +
  scale_x_log10() +
  theme_minimal() +
  labs(title = "Distribution of Total net_income by Firm Status",
       x = "net_income (log scale)",
       fill = "Status")
```

## Density Plot for total gross profit by status

```{r}
ggplot(american_banktupcy_dataset, aes(x = gross_profit,
                          fill = status_label)) +
  geom_density(alpha = 0.4) +
  scale_x_log10() +
  theme_minimal() +
  labs(title = "Distribution of Total gross profit by Firm Status",
       x = "gross_profit (log scale)",
       fill = "Status")
```

```{r}
ggplot(american_banktupcy_dataset, aes(x = status_label,
                          y = total_liabilities,
                          fill = status_label)) +
  geom_violin(alpha = 0.6) +
  scale_y_log10() +
  theme_minimal() +
  labs(title = "Total Liabilities by Firm Status")
```

```{r}
american_banktupcy_dataset$leverage_ratio <- american_banktupcy_dataset$total_liabilities / american_banktupcy_dataset$total_assets
american_banktupcy_dataset$profit_margin <- american_banktupcy_dataset$net_income / american_banktupcy_dataset$total_revenue

ggplot(american_banktupcy_dataset, aes(x = leverage_ratio,
                          y = profit_margin,
                          color = status_label)) +
  geom_point(alpha = 0.5) +
  theme_minimal() +
  labs(title = "Leverage vs Profitability by Firm Status",
       x = "Leverage Ratio (Liabilities / Assets)",
       y = "Profit Margin (Net Income / Revenue)")
```

```{r}
american_banktupcy_dataset$current_ratio <- american_banktupcy_dataset$current_assets / american_banktupcy_dataset$total_current_liabilities

ggplot(american_banktupcy_dataset, aes(x = total_long_term_debt,
                          y = current_ratio,
                          color = status_label)) +
  geom_point(alpha = 0.5) +
  scale_x_log10() +
  theme_minimal() +
  labs(title = "Debt vs Liquidity by Firm Status",
       x = "Total Long-Term Debt (log scale)",
       y = "Current Ratio")
```

```{r}
ggplot(american_banktupcy_dataset, aes(x = total_assets,
                          y = market_value,
                          color = status_label)) +
  geom_point(alpha = 0.5) +
  scale_x_log10() +
  scale_y_log10() +
  theme_minimal() +
  labs(title = "Market Value vs Total Assets by Firm Status")
```

```{r}
ggplot(american_banktupcy_dataset, aes(x = year,
                          y = leverage_ratio,
                          color = status_label)) +
  geom_jitter(alpha = 0.3) +
  theme_minimal() +
  labs(title = "Leverage Ratio Over Time by Firm Status")
```

```{r}
ggplot(american_banktupcy_dataset, aes(x = total_assets,
                          y = market_value,
                          color = status_label)) +
  geom_point(alpha = 0.5) +
  scale_x_log10() +
  scale_y_log10() +
  facet_wrap(~ year) +
  theme_minimal() +
  labs(title = "Market Value vs Total Assets by Year",
       x = "Total Assets (log scale)",
       y = "Market Value (log scale)",
       color = "Firm Status")
```

```{r}
ggplot(american_banktupcy_dataset %>% filter(year == 2018),
       aes(x = total_assets,
           y = market_value,
           color = status_label)) +
  geom_point(alpha = 0.5) +
  scale_x_log10() +
  scale_y_log10() +
  theme_minimal() +
  labs(title = "Market Value vs Total Assets (2018)",
       x = "Total Assets (log scale)",
       y = "Market Value (log scale)",
       color = "Firm Status")
```

```{r}
ggplot(american_banktupcy_dataset %>% filter(year == 2008),
       aes(x = total_assets,
           y = market_value,
           color = status_label)) +
  geom_point(alpha = 0.5) +
  scale_x_log10() +
  scale_y_log10() +
  theme_minimal() +
  labs(title = "Market Value vs Total Assets (2008)",
       x = "Total Assets (log scale)",
       y = "Market Value (log scale)",
       color = "Firm Status")
```

```{r}

ggplot(filter(american_banktupcy_dataset, status_label == "alive"),
       aes(x = year)) +
  geom_bar() +
  theme_minimal() +
  labs(title = "Active (Alive) Companies by Year",
       x = "Year",
       y = "Count of Companies")
```

```{r}
ggplot(filter(american_banktupcy_dataset, status_label == "failed"),
       aes(x = year)) +
  geom_bar() +
  theme_minimal() +
  labs(title = "Bankrupt (Failed) Companies by Year",
       x = "Year",
       y = "Count of Companies")
```

## Unique Companies that failed

```{r}
first_failure <- american_banktupcy_dataset %>%
  filter(status_label == "failed") %>%
  group_by(company_name) %>%
  summarise(first_bankrupt_year = min(year), .groups = "drop")

ggplot(first_failure, aes(x = first_bankrupt_year)) +
  geom_bar(fill = "firebrick") +
  theme_minimal() +
  labs(title = "First Bankruptcy Event by Year",
       x = "Year",
       y = "Number of Companies")
```

## Unique Companies alive

```{r}
first_failure <- american_banktupcy_dataset %>%
  filter(status_label == "alive") %>%
  group_by(company_name) %>%
  summarise(first_bankrupt_year = min(year), .groups = "drop")

ggplot(first_failure, aes(x = first_bankrupt_year)) +
  geom_bar(fill = "green") +
  theme_minimal() +
  labs(title = "First alive Event by Year",
       x = "Year",
       y = "Number of Companies")
```

## Splitting Data into training and test

```{r}
set.seed(123)

df <- american_banktupcy_dataset

# unique companies
companies <- df %>% distinct(company_name)

# sample 80% of companies into train
train_companies <- companies %>%
  sample_frac(0.80)

# split data by those companies
train_companies_df <- df %>% semi_join(train_companies, by = "company_name")
test_companies_df  <- df %>% anti_join(train_companies, by = "company_name")

# quick checks
cat("Train rows:", nrow(train_companies_df), "\n")
cat("Test rows:", nrow(test_companies_df), "\n")
cat("Train companies:", n_distinct(train_companies_df$company_name), "\n")
cat("Test companies:", n_distinct(test_companies_df$company_name), "\n")

# leakage check (should be 0)
length(intersect(unique(train_companies_df$company_name), unique(test_companies_df$company_name)))
```
```{r}

suppressPackageStartupMessages(library(tidyverse))

if (!requireNamespace("eulerr", quietly = TRUE)) install.packages("eulerr")
library(eulerr)

train_set <- unique(train_companies_df$company_name)
test_set  <- unique(test_companies_df$company_name)

overlap_n <- length(intersect(train_set, test_set))
train_n   <- length(train_set)
test_n    <- length(test_set)

fit <- eulerr::euler(c(
  "Train" = train_n,
  "Test"  = test_n,
  "Train&Test" = overlap_n
))

plot(
  fit,
  fills = c("steelblue2", "orange2"),
  labels = list(font = 2),
  quantities = TRUE,
  main = paste0(
    "Company-Level Split Overlap Check\n",
    "Train companies: ", train_n,
    " | Test companies: ", test_n,
    " | Overlap: ", overlap_n
  )
)

# -------------------------------------------------
# 2) Company-year heatmap (sample companies)
# -------------------------------------------------
set.seed(123)

sample_n_companies <- 15  # increase/decrease for readability on slides
sample_companies <- sample(unique(american_banktupcy_dataset$company_name),
                           size = sample_n_companies)

split_map <- tibble(
  company_name = c(train_set, test_set),
  split = c(rep("Train", length(train_set)), rep("Test", length(test_set)))
)

plot_df <- american_banktupcy_dataset %>%
  filter(company_name %in% sample_companies) %>%
  left_join(split_map, by = "company_name") %>%
  mutate(
    split = factor(split, levels = c("Train", "Test")),
    company_name = forcats::fct_reorder(company_name, year, .fun = min)
  )

ggplot(plot_df, aes(x = year, y = company_name, fill = split)) +
  geom_tile(color = "white", linewidth = 0.2) +
  labs(
    title = "Leakage-Safe Split Visualization (Sample Companies)",
    subtitle = paste0(
      "Each company belongs to ONLY one set. Company overlap = ", overlap_n,
      " (expected: 0)"
    ),
    x = "Year",
    y = "Company (sample)",
    fill = "Dataset"
  ) +
  theme_minimal(base_size = 16) +
  theme(panel.grid = element_blank())
```



## Transformation and scaling

```{r}
library(caret)

# keep IDs/outcome out of scaling
id_cols <- c("company_name", "status_label", "year")

numeric_cols <- names(train_companies_df)[sapply(train_companies_df, is.numeric)]
numeric_cols <- setdiff(numeric_cols, "year")  # don't scale year unless you want to

# Fit scaler on TRAIN only
pp <- preProcess(train_companies_df[, numeric_cols], method = c("center", "scale"))

# Apply to train and test
train_scaled <- train_companies_df
test_scaled  <- test_companies_df

train_scaled[, numeric_cols] <- predict(pp, train_companies_df[, numeric_cols])
test_scaled[, numeric_cols]  <- predict(pp, test_companies_df[, numeric_cols])

# sanity check
summary(train_scaled[, numeric_cols])
```

```{r}
ggplot(american_banktupcy_dataset, aes(x = gross_profit,
                          y = total_current_liabilities,
                          color = status_label)) +
  geom_jitter(alpha = 0.3) +
  theme_minimal() +
  labs(title = "Leverage Ratio Over Time by Firm Status")
```



## FEATURE ENGINEERING

# PROFITABILITY Ratio

```{r}
library(dplyr)

add_features <- function(df) {
  df %>%
    arrange(company_name, year) %>%
    group_by(company_name) %>%
    mutate(
      # Growth features (within company; computed only within this split)
      revenue_growth = (total_revenue - lag(total_revenue)) / lag(total_revenue),
      asset_growth   = (total_assets  - lag(total_assets))  / lag(total_assets),
      income_growth  = (net_income    - lag(net_income))    / lag(net_income),

      # Company-level features (computed within this split only)
      years_observed = n(),
      years_alive    = sum(status_label == "alive"),

      # “Eventually bankrupt” indicator (within this split only)
      eventually_bankrupt = ifelse(any(status_label == "failed"), 1, 0),

      # Survival window features (within this split only)
      first_year     = min(year),
      last_year      = max(year),
      survival_years = last_year - first_year + 1
    ) %>%
    ungroup() %>%
    mutate(
      # Profitability ratios (row-level; safe)
      profit_margin = ifelse(total_revenue == 0, NA, net_income / total_revenue),
      ebit_margin   = ifelse(total_revenue == 0, NA, ebit / total_revenue),
      ebitda_margin = ifelse(total_revenue == 0, NA, ebitda / total_revenue),
      gross_margin  = ifelse(total_revenue == 0, NA, gross_profit / total_revenue),

      # Leverage ratios (row-level; safe)
      leverage_ratio       = ifelse(total_assets == 0, NA, total_liabilities / total_assets),
      long_term_debt_ratio = ifelse(total_assets == 0, NA, total_long_term_debt / total_assets),
      debt_to_revenue      = ifelse(total_revenue == 0, NA, total_liabilities / total_revenue),

      # Working capital + flag (row-level; safe)
      working_capital = current_assets - total_current_liabilities,
      negative_income_flag = ifelse(net_income < 0, 1, 0)
    )
}
```

## Add Features to training and test

```{r}
train_companies_df <- add_features(train_companies_df)
test_companies_df  <- add_features(test_companies_df)
```

#Verify existance

```{r}
new_cols <- c(
  "profit_margin","ebit_margin","ebitda_margin","gross_margin",
  "leverage_ratio","long_term_debt_ratio","debt_to_revenue",
  "revenue_growth","asset_growth","income_growth",
  "working_capital","negative_income_flag",
  "years_observed","years_alive","eventually_bankrupt",
  "first_year","last_year","survival_years"
)

all(new_cols %in% names(train_companies_df))
all(new_cols %in% names(test_companies_df))
```

#No Overlap checks

```{r}
length(intersect(unique(train_companies_df$company_name),
                 unique(test_companies_df$company_name)))

library(tidyverse)
library(knitr)
library(kableExtra)

engineered_table <- tibble(
  Category = c(
    rep("Growth Features", 3),
    rep("Company-Level Features", 5),
    rep("Profitability Ratios", 4),
    rep("Leverage Ratios", 3),
    rep("Liquidity / Flags", 2)
  ),
  Engineered_Variable = c(
    "revenue_growth",
    "asset_growth",
    "income_growth",
    
    "years_observed",
    "years_alive",
    "eventually_bankrupt",
    "first_year",
    "survival_years",
    
    "profit_margin",
    "ebit_margin",
    "ebitda_margin",
    "gross_margin",
    
    "leverage_ratio",
    "long_term_debt_ratio",
    "debt_to_revenue",
    
    "working_capital",
    "negative_income_flag"
  )
)

engineered_table %>%
  kable(caption = "Engineered Variables Created for Clustering Analysis") %>%
  kable_styling(full_width = FALSE)
```

## Add Features to training and test

```{r}
train_companies_df <- add_features(train_companies_df)
test_companies_df  <- add_features(test_companies_df)
```

## Transformation and Scaling 
```{r}


suppressPackageStartupMessages({
  library(dplyr)
  library(caret)
})

id_cols <- c("company_name", "status_label")
numeric_cols <- train_companies_df %>%
  select(-all_of(id_cols)) %>%
  select(where(is.numeric)) %>%
  names()

# OPTIONAL: if you do NOT want to scale "year", remove it:
numeric_cols <- setdiff(numeric_cols, "year")

# 2) Fit preprocessing on training data only
# Choose methods:
# - center, scale = standardization
# - (optional) YeoJohnson handles skew and negatives safely
pp <- preProcess(
  train_companies_df[, numeric_cols],
  method = c("center", "scale")  # add "YeoJohnson" if you want power transform
  # method = c("YeoJohnson", "center", "scale")
)

# 3) Apply to train/test
train_scaled_numeric <- predict(pp, train_companies_df[, numeric_cols])
test_scaled_numeric  <- predict(pp, test_companies_df[, numeric_cols])

# 4) Rebuild scaled dataframes while keeping id + target columns
train_companies_scaled_df <- train_companies_df %>%
  select(all_of(id_cols)) %>%
  bind_cols(train_scaled_numeric)

test_companies_scaled_df <- test_companies_df %>%
  select(all_of(id_cols)) %>%
  bind_cols(test_scaled_numeric)

# 5) Quick validation checks
cat("Scaled train dimensions:", dim(train_companies_scaled_df), "\n")
cat("Scaled test dimensions:", dim(test_companies_scaled_df), "\n")

# Means should be ~0 and sds ~1 in TRAIN
train_means <- sapply(train_companies_scaled_df[, numeric_cols], mean, na.rm = TRUE)
train_sds   <- sapply(train_companies_scaled_df[, numeric_cols], sd, na.rm = TRUE)

cat("Train mean range (should be near 0):", range(train_means), "\n")
cat("Train sd range (should be near 1):", range(train_sds), "\n")
```


## Build a 2-means cluster model (k=2) and interpret the results.

```{r}
# Load packages
suppressPackageStartupMessages(library(tidyverse))

# Increase font size for ggplot2
theme_set(theme_gray(base_size = 18))


colors <- c("#1f77b4","#ff7f0e", "#2ca02c", "#d62728",
            "#9467bd","#8c564b", "#e377c2", "#7f7f7f",
            "#bcbd22", "#17becf")


cluster_vars <- c("leverage_ratio", "profit_margin")

clust_df <- train_companies_df %>%
  select(all_of(cluster_vars)) %>%
  drop_na()  # kmeans cannot handle NA

head(clust_df)

# OPTIONAL but recommended: cap extreme leverage (financial data can have huge outliers)
# Comment out if you don't want winsorization.
cap <- function(x, p = 0.01) {
  q <- quantile(x, probs = c(p, 1 - p), na.rm = TRUE)
  pmin(pmax(x, q[1]), q[2])
}

clust_df <- clust_df %>%
  mutate(
    leverage_ratio = cap(leverage_ratio, p = 0.01),
    profit_margin  = cap(profit_margin,  p = 0.01)
  )

# Scale (required for k-means distance)
clust_scaled <- scale(clust_df)


set.seed(123)
kmModel <- kmeans(clust_scaled, centers = 2, nstart = 25)
kmModel

# Add cluster assignments
clust_df$cluster <- as.factor(kmModel$cluster)
head(clust_df)


kmCenters_scaled <- as.data.frame(kmModel$centers)

scale_center <- attr(clust_scaled, "scaled:center")
scale_scale  <- attr(clust_scaled, "scaled:scale")

kmCenters <- sweep(kmCenters_scaled, 2, scale_scale, "*")
kmCenters <- sweep(kmCenters, 2, scale_center, "+")

colnames(kmCenters) <- cluster_vars


clust_df |>
  ggplot(aes(x = leverage_ratio, y = profit_margin)) +
  geom_point(aes(color = cluster, shape = cluster), cex = 4, alpha = 0.6) +
  geom_point(data = kmCenters,
             aes(x = leverage_ratio, y = profit_margin),
             color = "black", cex = 10, shape = 18) +
  labs(
    title = "k-means Clustering (k=2): Leverage vs Profitability",
    x = "Leverage Ratio (Total Liabilities / Total Assets)",
    y = "Profit Margin (Net Income / Total Revenue)"
  ) +
  theme(legend.position = "none") +
  scale_color_manual(values = colors)


clust_df %>%
  group_by(cluster) %>%
  summarise(
    leverage_mean = mean(leverage_ratio, na.rm = TRUE),
    leverage_sd   = sd(leverage_ratio, na.rm = TRUE),
    profit_mean   = mean(profit_margin, na.rm = TRUE),
    profit_sd     = sd(profit_margin, na.rm = TRUE),
    n = n()
  )


status_compare <- train_companies_df %>%
  select(all_of(cluster_vars), status_label) %>%
  drop_na() %>%
  mutate(
    leverage_ratio = cap(leverage_ratio, p = 0.01),
    profit_margin  = cap(profit_margin,  p = 0.01)
  )

# Attach cluster labels (same rows used for clustering)
status_compare$cluster <- clust_df$cluster

table(status_compare$cluster, status_compare$status_label)


set.seed(123)
WCSS <- sapply(1:5, function(k) kmeans(clust_scaled, k, nstart = 25)$tot.withinss)
WCSS

ggplot() +
  geom_point(aes(x = 1:5, y = WCSS), cex = 4) +
  geom_line(aes(x = 1:5, y = WCSS)) +
  labs(x = "Number of clusters (k)",
       y = "Within-cluster sum of squares (WCSS)")
```
## Apply Kmeans to Test

```{r}
# ---- Apply TRAIN k-means model to TEST data ----

cluster_vars <- c("leverage_ratio", "profit_margin")

# 1) Prep test data (same steps as train: select vars, drop NA, cap outliers)
test_clust_df <- test_companies_df %>%
  select(all_of(cluster_vars)) %>%
  drop_na() %>%
  mutate(
    leverage_ratio = cap(leverage_ratio, p = 0.01),
    profit_margin  = cap(profit_margin,  p = 0.01)
  )

# 2) Scale test using TRAIN center/scale (NO leakage)
test_scaled <- scale(test_clust_df, center = scale_center, scale = scale_scale)

# 3) Assign each test point to the nearest TRAIN cluster center (in scaled space)
# kmModel$centers are already in scaled space
assign_cluster <- function(X_scaled, centers_scaled) {
  # returns vector of cluster ids (1..k)
  apply(X_scaled, 1, function(x) {
    d <- apply(centers_scaled, 1, function(cntr) sum((x - cntr)^2))
    which.min(d)
  })
}

test_cluster_id <- assign_cluster(test_scaled, kmModel$centers)

# 4) Add cluster labels back to test dataframe
test_clust_df$cluster <- factor(test_cluster_id)

# If you want clusters attached to the full Test_companies_df (with NAs preserved),
# create a new column and fill only the rows that were used (complete cases on cluster_vars)
test_companies_df$cluster <- NA

test_rows_used <- complete.cases(test_companies_df[, cluster_vars])
# but we also dropped NA then capped; this matches complete.cases
test_companies_df$cluster[test_rows_used] <- as.character(test_clust_df$cluster)

# quick check
table(test_companies_df$cluster, useNA = "ifany")

plot_train <- clust_df %>% mutate(dataset = "Train")
plot_test  <- test_clust_df %>% mutate(dataset = "Test")

bind_rows(plot_train, plot_test) |>
  ggplot(aes(x = leverage_ratio, y = profit_margin)) +
  geom_point(aes(color = cluster, shape = dataset), cex = 4, alpha = 0.6) +
  geom_point(data = kmCenters,
             aes(x = leverage_ratio, y = profit_margin),
             color = "black", cex = 10, shape = 18) +
  labs(
    title = "Train vs Test: k-means Clusters (Train Fit, Test Assigned)",
    x = "Leverage Ratio",
    y = "Profit Margin"
  ) +
  scale_color_manual(values = colors)
```
```{r}
clust_df <- train_companies_df %>%
  select(all_of(cluster_vars), status_label) %>%
  drop_na() %>%
  mutate(
    leverage_ratio = cap(leverage_ratio, p = 0.01),
    profit_margin  = cap(profit_margin,  p = 0.01),
    status_label = factor(status_label)
  )
```



```{r}
cluster_vars <- c("leverage_ratio", "profit_margin")

# --- TRAIN (include year + status_label) ---
clust_df <- train_companies_df %>%
  select(all_of(cluster_vars), status_label, year) %>%
  drop_na() %>%
  mutate(
    leverage_ratio = cap(leverage_ratio, p = 0.01),
    profit_margin  = cap(profit_margin,  p = 0.01),
    status_label   = factor(status_label),
    year           = as.integer(year)
  )

# --- TEST (include year + status_label) ---
test_clust_df <- test_companies_df %>%
  select(all_of(cluster_vars), status_label, year) %>%
  drop_na() %>%
  mutate(
    leverage_ratio = cap(leverage_ratio, p = 0.01),
    profit_margin  = cap(profit_margin,  p = 0.01),
    status_label   = factor(status_label),
    year           = as.integer(year)
  )

# Apply TRAIN scaling to TEST (no leakage)
test_scaled <- scale(test_clust_df[, cluster_vars], center = scale_center, scale = scale_scale)

assign_cluster <- function(X_scaled, centers_scaled) {
  apply(X_scaled, 1, function(x) {
    d <- apply(centers_scaled, 1, function(cntr) sum((x - cntr)^2))
    which.min(d)
  })
}

test_clust_df$cluster <- factor(assign_cluster(test_scaled, kmModel$centers))

plot_train <- clust_df %>% mutate(dataset = "Train")
plot_test  <- test_clust_df %>% mutate(dataset = "Test")
combined_plot_data <- bind_rows(plot_train, plot_test)

# ---- CHOOSE YEAR HERE ----
year_to_plot <- 2018

plot_year <- combined_plot_data %>% filter(year == year_to_plot)

p <- ggplot(plot_year, aes(x = leverage_ratio, y = profit_margin)) +
  geom_point(aes(color = cluster, shape = dataset), size = 3.2, alpha = 0.6) +
  geom_point(
    data = plot_year %>% filter(status_label == "Failed"),
    aes(x = leverage_ratio, y = profit_margin),
    shape = 21, stroke = 1.1, color = "black", fill = NA, size = 3.2
  ) +
  geom_point(
    data = kmCenters,
    aes(x = leverage_ratio, y = profit_margin),
    color = "black", size = 6, shape = 18
  ) +
  labs(
    title = paste0("Train vs Test: k-means Clusters (Year ", year_to_plot, ")"),
    x = "Leverage Ratio",
    y = "Profit Margin",
    color = "Cluster",
    shape = "Dataset"
  ) +
  scale_color_manual(values = colors) +
  coord_cartesian(clip = "off") +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "right",
    plot.margin = margin(t = 10, r = 60, b = 10, l = 10),
    plot.title = element_text(hjust = 0.5)
  )

p
```



```{r}
# ---- Visualize TEST clusters ----
test_clust_df |>
  ggplot(aes(x = leverage_ratio, y = profit_margin)) +
  geom_point(aes(color = cluster, shape = cluster), cex = 4, alpha = 0.6) +
  # show TRAIN centers (same centers learned from train)
  geom_point(data = kmCenters,
             aes(x = leverage_ratio, y = profit_margin),
             color = "black", cex = 10, shape = 18) +
  labs(
    title = "TEST Set: Assigned k-means Clusters (k=2)",
    x = "Leverage Ratio (Total Liabilities / Total Assets)",
    y = "Profit Margin (Net Income / Total Revenue)"
  ) +
  theme(legend.position = "none") +
  scale_color_manual(values = colors)
```


# Cluster Kmeans
```{r}
library(dplyr)

# Choose clustering variables
cluster_vars <- c(
  "leverage_ratio",
  "long_term_debt_ratio",
  "profit_margin",
  "ebitda_margin",
  "debt_to_revenue",
  "working_capital"
)

# Remove rows with NA in clustering variables
train_cluster_data <- train_companies_scaled_df %>%
  select(all_of(cluster_vars)) %>%
  na.omit()

set.seed(123)

k2_model <- kmeans(train_cluster_data, centers = 2, nstart = 25)

# Attach clusters back to train dataframe
train_companies_scaled_df$cluster <- NA
train_companies_scaled_df$cluster[complete.cases(train_companies_scaled_df[, cluster_vars])] <- k2_model$cluster

table(train_companies_scaled_df$cluster)
# Function to assign nearest centroid
assign_cluster <- function(data, centers) {
  as.numeric(apply(data, 1, function(row) {
    which.min(colSums((t(centers) - row)^2))
  }))
}

# Prepare test data
test_cluster_data <- test_companies_scaled_df %>%
  select(all_of(cluster_vars))

# Assign clusters
test_companies_scaled_df$cluster <- assign_cluster(
  as.matrix(test_cluster_data),
  k2_model$centers
)

table(test_companies_scaled_df$cluster)
library(ggplot2)

ggplot(train_companies_scaled_df,
       aes(x = leverage_ratio,
           y = profit_margin,
           color = factor(cluster))) +
  geom_point(alpha = 0.4) +
  labs(
    title = "K-Means Clustering (k = 2) on Scaled Financial Structure",
    x = "Leverage Ratio (Scaled)",
    y = "Profit Margin (Scaled)",
    color = "Cluster"
  ) +
  theme_minimal(base_size = 14)
```
```{r}
table(train_companies_scaled_df$cluster,
      train_companies_scaled_df$status_label)

prop.table(
  table(train_companies_scaled_df$cluster,
        train_companies_scaled_df$status_label),
  margin = 1
)
```


# Cluster Revenue and operating expenses

```{r}


suppressPackageStartupMessages(library(tidyverse))

theme_set(theme_gray(base_size = 18))

colors <- c("#1f77b4","#ff7f0e", "#2ca02c", "#d62728",
            "#9467bd","#8c564b", "#e377c2", "#7f7f7f",
            "#bcbd22", "#17becf")


cluster_vars <- c("total_revenue", "total_operating_expenses")

clust_df <- train_companies_df %>%
  select(all_of(cluster_vars)) %>%
  drop_na()  # kmeans cannot handle NA

head(clust_df)


clust_scaled <- scale(clust_df)


set.seed(123)
kmModel <- kmeans(clust_scaled, centers = 2, nstart = 25)
kmModel

# Add cluster assignments
clust_df$cluster <- as.factor(kmModel$cluster)
head(clust_df)


kmCenters_scaled <- as.data.frame(kmModel$centers)

scale_center <- attr(clust_scaled, "scaled:center")
scale_scale  <- attr(clust_scaled, "scaled:scale")

kmCenters <- sweep(kmCenters_scaled, 2, scale_scale, "*")
kmCenters <- sweep(kmCenters, 2, scale_center, "+")

colnames(kmCenters) <- cluster_vars


clust_df |>
  ggplot(aes(x = total_revenue, y = total_operating_expenses)) +
  geom_point(aes(color = cluster, shape = cluster), cex = 4, alpha = 0.6) +
  geom_point(data = kmCenters,
             aes(x = total_revenue, y = total_operating_expenses),
             color = "black", cex = 10, shape = 18) +
  labs(
    title = "k-means Clustering (k=2): Revenue vs Operating Expenses",
    x = "Total Revenue",
    y = "Total Operating Expenses"
  ) +
  theme(legend.position = "none") +
  scale_color_manual(values = colors)


clust_df %>%
  group_by(cluster) %>%
  summarise(
    revenue_mean  = mean(total_revenue),
    revenue_sd    = sd(total_revenue),
    expense_mean  = mean(total_operating_expenses),
    expense_sd    = sd(total_operating_expenses),
    n = n()
  )


status_compare <- train_companies_df %>%
  select(all_of(cluster_vars), status_label) %>%
  drop_na()

# Attach cluster labels (same rows used for clustering)
status_compare$cluster <- clust_df$cluster

table(status_compare$cluster, status_compare$status_label)


set.seed(123)
WCSS <- sapply(1:5, function(k) kmeans(clust_scaled, k, nstart = 25)$tot.withinss)
WCSS

ggplot() +
  geom_point(aes(x = 1:5, y = WCSS), cex = 4) +
  geom_line(aes(x = 1:5, y = WCSS)) +
  labs(x = "Number of clusters (k)",
       y = "Within-cluster sum of squares (WCSS)")
```

```{r}
prop.table(
  table(status_compare$cluster,
        status_compare$status_label),
  margin = 1
)
```

## Confusion Matrix

```{r}

suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(caret))

set.seed(123)


pos_class <- "failed"  # positive class
cluster_vars <- c("leverage_ratio", "profit_margin")  # your chosen pair


# 1) Prepare TRAIN data for k-means (drop NA, keep only vars)

train_km <- train_companies_df %>%
  mutate(status_label = stringr::str_trim(stringr::str_to_lower(status_label))) %>%
  select(all_of(cluster_vars), status_label) %>%
  drop_na()

X_train <- train_km %>%
  select(all_of(cluster_vars))

# Scale using TRAIN statistics ONLY (no leakage)
train_center <- colMeans(X_train)
train_scale  <- apply(X_train, 2, sd)

X_train_scaled <- scale(X_train, center = train_center, scale = train_scale)

# Fit k-means (k=2)
kmModel <- kmeans(X_train_scaled, centers = 2, nstart = 25)

# Add cluster labels back to TRAIN rows used
train_km$cluster <- kmModel$cluster


# 2) Map cluster 

cluster_rates <- train_km %>%
  group_by(cluster) %>%
  summarise(
    n = n(),
    failed_rate = mean(status_label == pos_class),
    .groups = "drop"
  )

# Cluster with highest failed_rate becomes "failed"
failed_cluster <- cluster_rates %>%
  slice_max(failed_rate, n = 1, with_ties = FALSE) %>%
  pull(cluster)

cluster_map <- tibble(
  cluster = sort(unique(train_km$cluster)),
  predicted_label = ifelse(cluster == failed_cluster, "failed", "alive")
)

cluster_map

# This table tells you, e.g. cluster 1 -> "alive", cluster 2 -> "failed" (or vice versa)


# 3) Prepare TEST data (same vars), scale with TRAIN params, assign nearest centroid

test_km <- test_companies_df %>%
  mutate(status_label = stringr::str_trim(stringr::str_to_lower(status_label))) %>%
  select(all_of(cluster_vars), status_label) %>%
  drop_na()

X_test <- test_km %>%
  select(all_of(cluster_vars))

X_test_scaled <- scale(X_test, center = train_center, scale = train_scale)

# Centroids from kmeans are in SCALED space
centers <- kmModel$centers  # 2 x p matrix

# Assign each test row to nearest centroid (Euclidean distance)
assign_cluster <- function(X, centers_mat) {
  # X: matrix n x p
  # centers_mat: k x p
  dists <- sapply(1:nrow(centers_mat), function(k) {
    rowSums((X - matrix(centers_mat[k, ], nrow = nrow(X), ncol = ncol(X), byrow = TRUE))^2)
  })
  max.col(-dists)  # returns index of smallest distance
}

test_clusters <- assign_cluster(as.matrix(X_test_scaled), centers)
test_km$cluster <- test_clusters

# Attach predicted label using mapping from TRAIN
test_pred <- test_km %>%
  left_join(cluster_map, by = "cluster") %>%
  mutate(
    status_label = factor(status_label, levels = c("alive", "failed")),
    predicted_label = factor(predicted_label, levels = c("alive", "failed"))
  )


# 4) Confusion matrix + metrics on TEST

conf_matrix <- table(
  Actual = test_pred$status_label,
  Predicted = test_pred$predicted_label
)
conf_matrix

TN <- conf_matrix["alive", "alive"]
FP <- conf_matrix["alive", "failed"]
FN <- conf_matrix["failed", "alive"]
TP <- conf_matrix["failed", "failed"]

accuracy  <- (TP + TN) / sum(conf_matrix)
precision <- ifelse((TP + FP) == 0, NA, TP / (TP + FP))      # PPV
recall    <- ifelse((TP + FN) == 0, NA, TP / (TP + FN))      # Sensitivity
f1        <- ifelse(is.na(precision) | is.na(recall) | (precision + recall) == 0,
                    NA,
                    2 * (precision * recall) / (precision + recall))

metrics <- tibble(
  Metric = c("Accuracy", "Precision (PPV)", "Recall (Sensitivity)", "F1-score"),
  Value  = c(accuracy, precision, recall, f1)
)
metrics

# caret output (nice for reporting)
cm <- confusionMatrix(
  data = test_pred$predicted_label,
  reference = test_pred$status_label,
  positive = pos_class
)
cm
```

```{r}
write.csv(test_companies_df, "test_companies_df.csv", row.names = FALSE)
write.csv(train_companies_df, "train_companies_df.csv", row.names = FALSE)
```
